---
title: 'MATH 427: Linear Regression'
author: Eric Friedlander <br> [Much of the content in these slides have been adapted from Abhishek Chakraborty at Lawrence University]{.smaller}
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

```{r setup}
#| include: false
library(tidyverse)
library(ISLR2)
library(GGally)
library(gridExtra)
library(kableExtra)
library(plotly)
```

## Concept Check

What do we call these:

-   $\mathbf{X}$
-   $Y$

What is our goal?

## Concept Check

-   Features: $\mathbf{X}$
-   Target: $Y$
-   Goal: Predict $Y$ using $\mathbf{X}$

. . .

-   What do we call this process if $Y$ is numerical? What about categorical?

## Concept Check

-   What do we call this process if $Y$ is numerical? What about categorical?
    +   Numerical: regression
    +   Categorical: classification
    
. . .

-   What is the difference between a training and a test set?
-   How do we evaluate the performance of a regression model?


## A Familiar Supervised Learning Model {.smaller}

- Assume relationship between $\mathbf{X}$ and $Y$ is:
$$Y=f(\mathbf{X}) + \epsilon$$
where $\epsilon$ is a **random** error term (includes measurement error, other discrepancies) independent of $\mathbf{X}$ and has mean zero.
- **Objective**: To approximate/estimate $f(\mathbf{X})$
- **Linear Regression**: assume that $f(\mathbf{X})$ is a linear function of $\mathbf{X}$
  + For $p=1$: $f(\mathbf{X}) = \beta_0 + \beta_1 X_1$
  + For $p > 1$: $f(\mathbf{X}) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$

## Linear Regression {.smaller}

Suppose the CEO of a restaurant franchise is considering opening new outlets in different cities. They would like to expand their business to cities that give them higher profits with the assumption that highly populated cities will probably yield higher profits.

They have data on the population (in 100,000) and profit (in $1,000) at 97 cities where they currently have outlets.

```{r}
outlets <- readRDS("../data/outlets.rds")   # load dataset
head(outlets) |> kable() # first six observations of the dataset
```

## Linear Regression

- **Objective**: choose "best" $\beta_0$ and $\beta_1$ if we assume
$$\text{profit} = \beta_0 + \beta_1\times\text{population}$$
- Once this is done, we can:
  + predict the profit for a new city with a given population
  + understand the relationship between `population` and `profit` better

## Outlet EDA

```{r}
outlets |> ggpairs()
```

## Linear Regression: Estimating Parameters {.smaller}

::: {.incremental}
- Suppose we have $\hat{\beta}_0$ and $\hat{\beta}_1$
- Observed response: $y_i$ for $i=1,\ldots,n$
- Predicted response: $\hat{y}_i$ for $i=1, \ldots, n$
- Residual: $e_i = \hat{y}_i - y_i$ for $i=1, \ldots, n$
- **Mean Squared Error (MSE)**: $MSE =\dfrac{e^2_1+e^2_2+\ldots+e^2_n}{n}$  also known as the **loss/cost function**
- **GOAL**: Find $\hat{\beta}_0$ and $\hat{\beta}_1$ which minimizes $MSE$
:::


## Gradient Descent Algorithm

- How do we minimize the $MSE$?
  + Can be done "analytically" but most ML algorithms can't be fit that way
- **Today**: popular optimization algorithm called **gradient descent**.
- **NOTE**: Gradient Descent is not a machine learning technique. It is an optimization technique that helps to fit machine learning models.

## Gradient Descent Algorithm {.smaller}

- Think of the MSE as a function of $\hat{\beta}_0$ and $\hat{\beta}_1$
  + $\text{MSE} = \frac{\sum (y_i - \hat{\beta}_0 - \hat{\beta}_1)}{n}$

```{r}
#| echo: FALSE

centerb0 <- -4
centerb1 <- 1
b0_width <- 500
b1_width <- 50

b0 <- seq(from = -b0_width + centerb0, to = b0_width + centerb0, by = b0_width/100)
b1 <- seq(from = -b1_width + centerb1, to = b1_width + centerb1, by = b1_width/100)

grid <-  expand.grid(b0s=b0, b1s=b1)

axy <- list(
  range = c(min(b0), max(b0)),
  title.text = TeX("\\beta_0")
)

axx <- list(
  range = c(min(b1),max(b1)),
  title.text = TeX("\\beta_1")
)



grid <- tibble(grid) |>
  rowwise() |>
  mutate(MSE = mean((outlets$profit - b0s - b1s*outlets$population)^2))
min_val <- min(grid$MSE)
max_val <- min_val+.04*diff(range(grid$MSE))
axz <- list(
  range = c(min_val,max_val),
  title.text = "MSE"
)

grid_mat <- xtabs(MSE ~ b0s + b1s, data = grid)

plot_ly(x = ~b1, y = ~b0) |>
  add_surface(z = ~grid_mat, cmin=min_val, cmax=max_val, alpha = 0.7) |> 
  layout(scene = list(xaxis=axx,yaxis=axy, zaxis=axz)) |> 
  config(mathjax = 'cdn')
```

## Gradient Descent Algorithm {.smaller}

```{r}
#| echo: FALSE

# create data to plot

x <- seq(-5, 5, by = .05)

y <- x^2 + 3

df <- data.frame(x, y)

step <- 5

step_size <- .2

for(i in seq_len(18)) {

  next_step <- max(step) + round(diff(range(max(step), which.min(df$y))) * step_size, 0)

  step <- c(step, next_step)

  next

}

steps <- df[step, ] %>%

  mutate(x2 = lag(x), y2 = lag(y)) %>%

  dplyr::slice(1:18)

# plot

ggplot(df, aes(x, y)) +
  geom_line(linewidth = 1.5, alpha = .5) +
  geom_point(aes(x=steps$x[1], y = steps$y[1]), size = 3, shape = 21, fill = "blue", alpha = .5) +
  theme_classic() +
  scale_y_continuous("Cost/Loss function (MSE)", limits = c(0, 30)) +
  xlab("parameter to estimate")

```

Updates to the variable:
$$
\begin{aligned}
\text{new value of variable} &= \text{old value of variable}\\
&\qquad- \text{step size} \times \text{gradient of function with respect to variable}
\end{aligned}
$$

## Gradient Descent Algorithm

```{r}
#| echo: FALSE

# plot

ggplot(df, aes(x, y)) +

  geom_line(linewidth = 1.5, alpha = .5) +

  theme_classic() +

  scale_y_continuous("Cost/Loss function (MSE)", limits = c(0, 30)) +

  xlab("parameter to estiamte") +

  geom_segment(data = df[c(5, which.min(df$y)), ], aes(x = x, y = y, xend = x, yend = -Inf), lty = "dashed") +

  geom_point(data = filter(df, y == min(y)), aes(x, y), size = 4, shape = 21, fill = "yellow") +

  geom_point(data = steps, aes(x, y), size = 3, shape = 21, fill = "blue", alpha = .5) +

  geom_curve(data = steps[-1,], aes(x = x, y = y, xend = x2, yend = y2), curvature = 1, lty = "dotted") +

  theme(

    axis.ticks = element_blank(),

    axis.text = element_blank()

  ) +

  annotate("text", x = df[5, "x"], y = 1, label = "initial value", hjust = -0.1, vjust = .8) +

  annotate("text", x = df[which.min(df$y), "x"], y = 1, label = "minimium", hjust = -0.1, vjust = .8) +

  annotate("text", x = df[5, "x"], y = df[5, "y"], label = "learning rate (step size)", hjust = -.8, vjust = 0)

```

## Gradient Descent Algorithm {.smaller}

```{r}
#| echo: FALSE
#| layout-ncol: 2
#| fig-subcap: 
#|   - "Step size too big"
#|   - "Step size too small"


# create too small of a learning rate

step <- 5

step_size <- .05

for(i in seq_len(10)) {

  next_step <- max(step) + round(diff(range(max(step), which.min(df$y))) * step_size, 0)

  step <- c(step, next_step)

  next

}

too_small <- df[step, ] %>%

  mutate(x2 = lag(x), y2 = lag(y))

# plot

ggplot(df, aes(x, y)) +

  geom_line(size = 1.5, alpha = .5) +

  theme_classic() +

  scale_y_continuous("Cost/Loss function of the variable", limits = c(0, 30)) +

  xlab("variable") +

  geom_segment(data = too_small[1, ], aes(x = x, y = y, xend = x, yend = -Inf), lty = "dashed") +

  geom_point(data = too_small, aes(x, y), size = 3, shape = 21, fill = "blue", alpha = .5) +

  geom_curve(data = too_small[-1, ], aes(x = x, y = y, xend = x2, yend = y2), curvature = 1, lty = "dotted") +

  theme(

    axis.ticks = element_blank(),

    axis.text = element_blank()

  ) +

  annotate("text", x = df[5, "x"], y = 1, label = "initial value", hjust = -0.1, vjust = .8)

# create too large of a learning rate

too_large <- df[round(which.min(df$y) * (1 + c(-.9, -.6, -.2, .3)), 0), ] %>%

  mutate(x2 = lag(x), y2 = lag(y))

# plot

ggplot(df, aes(x, y)) +

  geom_line(size = 1.5, alpha = .5) +

  theme_classic() +

  scale_y_continuous("Cost/Loss function of the variable", limits = c(0, 30)) +

  xlab("variable") +

  geom_segment(data = too_large[1, ], aes(x = x, y = y, xend = x, yend = -Inf), lty = "dashed") +

  geom_point(data = too_large, aes(x, y), size = 3, shape = 21, fill = "blue", alpha = .5) +

  geom_curve(data = too_large[-1, ], aes(x = x, y = y, xend = x2, yend = y2), curvature = 1, lty = "dotted") +

  theme(

    axis.ticks = element_blank(),

    axis.text = element_blank()

  ) +

  annotate("text", x = too_large[1, "x"], y = 1, label = "initial value", hjust = -0.1, vjust = .8)
```

## Gradient Descent for Linear Regression {.smaller}

- **Objective**: We want to find $\hat{\beta}_0$ and $\hat{\beta}_1$ which minimizes
- $$MSE = \dfrac{e^2_1+e^2_2+\ldots+e^2_n}{n} = \dfrac{(\hat{y}_1 - y_1)^2 +  (\hat{y}_2 - y_2)^2 + \ldots + (\hat{y}_n - y_n)^2}{n}$$
- $$MSE = \dfrac{(\hat{\beta}_0 + \hat{\beta}_1 \ x_1 - y_1)^2 +  (\hat{\beta}_0 + \hat{\beta}_1 \ x_2 - y_2)^2 + \ldots + (\hat{\beta}_0 + \hat{\beta}_1 \ x_n - y_n)^2}{n}$$
- $$MSE = \displaystyle \dfrac{1}{n}\sum_{i=1}^{n} (\hat{\beta}_0 + \hat{\beta}_1 \ x_i - y_i)^2 = \displaystyle \dfrac{1}{n}\sum_{i=1}^{n} (\hat{y}_i - y_i)^2 = \displaystyle \dfrac{1}{n}\sum_{i=1}^{n} (e_i)^2$$

## Gradient Descent for Linear Regression {.smaller}

- To compute gradient, need partial derivatives of $MSE$ with respect to $\hat{\beta}_0$ and $\hat{\beta}_1$.
- $$\text{gradient of MSE with respect to} \ \hat{\beta}_0 = \dfrac{2}{n} \displaystyle \sum_{i=1}^{n} (\hat{\beta}_0 + \hat{\beta}_1 \ x_i - y_i) = \dfrac{2}{n} \displaystyle \sum_{i=1}^{n} (\hat{y}_i - y_i)$$
- $$\text{gradient of MSE with respect to} \ \hat{\beta}_1 = \dfrac{2}{n} \displaystyle \sum_{i=1}^{n} x_i (\hat{\beta}_0 + \hat{\beta}_1 \ x_i - y_i) = \dfrac{2}{n} \displaystyle \sum_{i=1}^{n} x_i (\hat{y}_i - y_i)$$

## Gradient Descent for Linear Regression {.smaller}

- Gradient descent update:
  + For $\hat{\beta}_0$:
    $$
    \begin{aligned}
    \hat{\beta}_0 \ \text{(new)} 
    &= \hat{\beta}_0 \ \text{(old)}- \bigg(\text{step size} \times \text{derivative w.r.t} \ \hat{\beta}_0\bigg)\\ 
    &= \hat{\beta}_0 \ \text{(old)}- \bigg(\text{step size} \times \dfrac{2}{n} \displaystyle \sum_{i=1}^{n} (\hat{y}_i - y_i)\bigg)
    \end{aligned}
    $$
  + For $\hat{\beta}_1$:
  $$
  \begin{aligned}
  \hat{\beta}_1  \ \text{(new)} &= \hat{\beta}_1  \ \text{(old)} - \bigg(\text{step size} \times \text{derivative w.r.t} \ \hat{\beta}_1 \bigg)\\
  &= \hat{\beta}_1  \ \text{(old)} - \bigg(\text{step size} \times \dfrac{2}{n} \displaystyle \sum_{i=1}^{n} x_i (\hat{y}_i - y_i)\bigg)
  \end{aligned}
  $$


## Linear Regression in R {.smaller}

::::{.columns}
:::{.column}
```{r}
outlets_model <- lm(profit ~ population, data = outlets)
outlets_model
```
:::

:::{.column}

This corresponds to the model:

$$
\begin{aligned}
\text{Profit}  &= -3.90 + 1.19\times\text{Population}\\
\hat{Y}_i &= -3.90 + 1.19X_i
\end{aligned}
$$
i.e. $\hat{\beta}_0 = -3.90$ and $\hat{\beta}_1 = 1.19$
:::
::::

## Linear Regression in R

```{r}

ggplot(data = outlets) +
  geom_point(mapping = aes(x = population, y = profit)) +    # create scatterplot
  geom_smooth(mapping = aes(x = population, y = profit), 
              method = "lm", se = FALSE)   # add the regression line
```

## Linear Regression in R: Prediction

```{r}
predict(outlets_model, newdata = data.frame(population = 17))
```

Note: New data must be a data frame with the same columns names as the training data
