---
title: 'MATH 427: Preprocessing, Missing Data, and Resampling'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---



## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(janitor) # for contingency tables
library(ISLR2)
library(AmesHousing)

tidymodels_prefer()

set.seed(427)
```

# Repeated Cross-Validation

## Repeated K-Fold Cross-Validation (CV) {.smaller}

:::{.incremental}
-   Partition your data into $K$ randomly selected non-overlapping "folds"
    +   Folds don't overlap and every training observation is in one fold
    +   Each fold contained $1/K$ of the training data
-   Looping through the folds $k = 1, \ldots, K$:
    +   Treat fold $k$ as the assessment set
    +   Treat all folds except for $k$ as the analysis set
    +   Fit model to analysis set (use whole modeling workflow)
    +   Compute error metrics on assessment set
-   After loop, you will have $K$ copies of each error metrics
-   Average them together to get performance estimate
-   Can also look at distribution of performance metrics
-   Repeat this process from the beginning selected $K$ different randomly chosen folds
-   If we want more accurate estimate you can perform repeated CV with different randomly chosen folds

:::

# CV Workflow in R

## Data: Ames Housing Prices {.smaller}

A data set from [De Cock (2011)](https://jse.amstat.org/v19n3/decock.pdf) has 82 fields were recorded for 2,930 properties in Ames IA. This version is copies from the `AmesHousing` package but does not include a few quality columns that appear to be outcomes rather than predictors.

Goal: Predict `Sale_Price`.

```{r}
ames |> glimpse()
```

## Initial Data Split

```{r}
set.seed(427)

data_split <- initial_split(ames, strata = "Sale_Price")
ames_train <- training(data_split)
ames_test  <- testing(data_split)
```

## Define Folds

```{r}
ames_folds <- vfold_cv(ames_train, v = 10, repeats = 10)
ames_folds
```

## Define Model(s)

```{r}
lm_model <- linear_reg() |>
  set_engine('lm')

knn5_model <- nearest_neighbor(neighbors = 5) |>
  set_engine("kknn") |>
  set_mode("regression")

knn10_model <- nearest_neighbor(neighbors = 10) |>
  set_engine("kknn") |>
  set_mode("regression")
```

## Define Preprocessing: Linear regression

```{r}
lm_preproc <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_dummy(all_nominal_predictors()) |> # Convert categorical data into dummy variables
  step_zv(all_predictors()) |> # remove zero-variance predictors (i.e. predictors with one value)
  step_corr(all_predictors(), threshold = 0.5) |> # remove highly correlated predictors
  step_lincomb(all_predictors()) # remove variables that have exact linear combinations
```

## Define Preprocessing: KNN

```{r}
knn_preproc <- recipe(Sale_Price ~ ., data = ames_train) |> # only uses ames_train for data types
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |> # Convert categorical data into dummy variables
  step_zv(all_predictors()) |> # remove zero-variance predictors (i.e. predictors with one value)
  step_normalize(all_predictors())
```

## Define Workflows

```{r}
lm_wf <- workflow() |> add_model(lm_model) |> add_recipe(lm_preproc)
knn5_wf <- workflow() |> add_model(knn5_model) |> add_recipe(knn_preproc)
knn10_wf <- workflow() |> add_model(knn10_model) |> add_recipe(knn_preproc)
```

## Define Metrics

```{r}
ames_metrics <- metric_set(rmse, rsq)
```

## Fit and Assess Models

```{r}
lm_results <- lm_wf |> fit_resamples(resamples = ames_folds, metrics = ames_metrics)
knn5_results <- knn5_wf |> fit_resamples(resamples = ames_folds, metrics = ames_metrics)
knn10_results <- knn10_wf |> fit_resamples(resamples = ames_folds, metrics = ames_metrics)
```


## Collecting Metrics {.smaller}

```{r}
collect_metrics(lm_results) |> kable()
collect_metrics(knn5_results) |> kable()
collect_metrics(knn10_results) |> kable()
```

## Final Model

- After choosing best model/workflow, fit on full training set and assess on test set

```{r}
final_fit <- lm_wf |> fit(data = ames_train)
final_fit_perf <- final_fit |> 
  predict(new_data = ames_test) |> 
  bind_cols(ames_test) |> 
  ames_metrics(truth = Sale_Price, estimate = .pred)

final_fit_perf |> kable()
```

## Extracting CV Metrics

-   `collect_metrics` averages over all 100 models
-   set `summarize = FALSE` to get all the individual errors

```{r}
collect_metrics(lm_results, summarize = FALSE) |> head() |> kable()
```

## Combining CV Metrics

```{r}
all_metrics <- bind_cols(method = "lm", collect_metrics(lm_results, summarize = FALSE)) |>
  bind_rows(bind_cols(method = "knn5", collect_metrics(knn5_results, summarize = FALSE))) |>
  bind_rows(bind_cols(method = "knn10", collect_metrics(knn10_results, summarize = FALSE)))
```

## Visualizing CV Metrics

:::: columns
::: column
```{r}
all_metrics |> 
  filter(.metric == "rmse") |> 
  ggplot(aes(x = method, y = .estimate)) +
  geom_boxplot() +
  geom_hline(yintercept=final_fit_perf |> filter(.metric == "rmse") |> pull(.estimate), color = "red")
```
:::

::: column
```{r}
all_metrics |> 
  filter(.metric == "rsq") |> 
  ggplot(aes(x = method, y = .estimate)) +
  geom_boxplot() +
  geom_hline(yintercept=final_fit_perf |> filter(.metric == "rsq") |> pull(.estimate), color = "red")
```
:::
::::

```{r}
#| echo: FALSE

ames <- ames |> 
  mutate(Overall_Cond = factor(Overall_Cond, levels = sample(levels(Overall_Cond))))
```


# Pre-processing


## Data: Different Ames Housing Prices {.smaller}

A data set from [De Cock (2011)](https://jse.amstat.org/v19n3/decock.pdf) has 82 fields were recorded for 2,930 properties in Ames IA. This version is the original (kind of) from the `AmesHousing` package.

Goal: Predict `Sale_Price`.

```{r}
ames <- make_ames()
ames |> glimpse()
```

```{r}
#| echo: FALSE
ames <- make_ames() |> 
  mutate(Overall_Cond = factor(Overall_Cond, levels = sample(levels(Overall_Cond))))
```

## Today

Well cover some common pre-processing tasks:

- Dealing with zero-variance (zv) and/or near-zero variance (nzv) variables
- Imputing missing entries
- Label encoding ordinal categorical variables
- Standardizing (centering and scaling) numeric predictors
- Lumping predictors
- One-hot/dummy encoding categorical predictor

## Pre-Split Cleaning

- Before you split your data: make sure data is in correct format
- This may mean different things for different data sets
- Common examples:
    + Fixing names of columns
    + Ensure all variable types are correct
    + Ensure all factor levels are correct and in order (if applicable)
    + Remove any variables that are not important (or harmful) to your analysis
    + Ensure missing values are coded as such (i.e. as `NA` instead of 0 or -1 or "missing")
    + Filling in missing values where you know what the answer should be (i.e. if a missing value really means 0 instead of missing)

## Example: Factor Levels in Wrong Order

```{r}
ames |> pull(Overall_Cond) |> levels()
```

## Re-Factoring

```{r}
ames <- ames |> 
  mutate(Overall_Cond = factor(Overall_Cond, levels = c("Very_Poor", "Poor", 
                                                        "Fair", "Below_Average",
                                                        "Average", "Above_Average", 
                                                        "Good", "Very_Good",
                                                        "Excellent", "Very_Excellent")))
ames |> pull(Overall_Cond) |> levels()
```

## Initial Data Split

```{r}
data_split <- initial_split(ames, strata = "Sale_Price")
ames_train <- training(data_split)
ames_test  <- testing(data_split)
```

## Zero-Variance (zv) and/or Near-Zero Variance (nzv) Variables {.smaller}

Heuristic for detecting near-zero variance features is:

- The fraction of unique values over the sample size is low (say ≤ 10%).
- The ratio of the frequency of the most prevalent value to the frequency of the second most prevalent
value is large (say ≥ 20%).

```{r}
library(caret)
nearZeroVar(ames_train, saveMetrics = TRUE) |> kable()
```

## Missing Data

-   Many times, you can't just drop missing data
-   Even if you can, dropping missing values can generate *biased* data/models
-   Sometimes missing data gives you more information
-   Types of missing data:
    +   Missing completely at random (MCAR): there is no pattern to your missing values
    +   Missing at random (MAR): missing values are dependent on other values in the data set
    +   Missing not at random (MNAR): missing values are dependent on the value that is missing
-   Structured missingness (SM):  when the missingness of certain values are depends on one another, regardless of whether the missing values are MCAR, MAR, or MNAR

## MCAR: Examples

-   Sensor data: occasionally sensors break so you're missing data randomly
-   Survey data: sometimes people just randomly skip questions
-   Survey data: customers are randomly given 5 questions from a bank of 100 questions

## MAR: Examples

-   Men are less likely to respond to surveys about depression
-   Medical study: patients who miss follow-up appointments are more likely to be young
-   Survey responses: ESL respondents may be more likely to skip certain questions that are difficult to interpret (only MAR if you know they are ESL)
-   Measure of student performance: students who score lower are more likely to skip questions

## MNAR: Examples

-   Survey on income: respondent may be less likely to report their income if they are poor
-   Survey about political beliefs: respondent may be more likely to skip questions when their answer is perceived as undesirable
-   Customer satisfaction: only customers who feel strongly respond
-   Medical study: patients refuse to report unhealthy habits

## Structurally Missing: Examples

-   Health survey: all questions related to pregnancy are left blank by males
-   Bank data set: combination of home, auto, and credit cards... not all customer have all three so have missing data in certain portions
-   Survey: many respondents by stop the survey early so all questions after a certain point are missing
-   Netflix: customers may only watch similar movies and TV shows

## Remedies for Missing Data

-   Lot of complicated ways that you can read about
-   Can drop column of too much of the data is missing
-   Imputing:
    +  `step_impute_median`: used for numeric (especially discrete) variables
    +  `step_impute_mean`: used for numeric variables
    +  `step_impute_knn`: used for both numeric and categorical variables (computationally expensive)
    +  `step_impute_mode`: used for nominal (having no order) categorical variable