---
title: 'MATH 427: Preprocessing, Missing Data, and Resampling'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---



## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(janitor) # for contingency tables
library(ISLR2)
library(AmesHousing)

tidymodels_prefer()

ames <- make_ames()

set.seed(427)
```

```{r}
#| echo: FALSE

ames <- ames |> 
  mutate(Overall_Cond = factor(Overall_Cond, levels = sample(levels(Overall_Cond))))
```


# Pre-processing

## Data: Ames Housing Prices {.smaller}

A data set from [De Cock (2011)](https://jse.amstat.org/v19n3/decock.pdf) has 82 fields were recorded for 2,930 properties in Ames IA. This version is the original from the `AmesHousing` package.

Goal: Predict `Sale_Price`.

```{r}
ames |> glimpse()
```
## Today

Well cover some common pre-processing tasks:

- Dealing with zero-variance (zv) and/or near-zero variance (nzv) variables
- Imputing missing entries
- Label encoding ordinal categorical variables
- Standardizing (centering and scaling) numeric predictors
- Lumping predictors
- One-hot/dummy encoding categorical predictor

## Pre-Split Cleaning

- Before you split your data: make sure data is in correct format
- This may mean different things for different data sets
- Common examples:
  + Fixing names of columns
  + Ensure all variable types are correct
  + Ensure all factor levels are correct and in order (if applicable)
  + Remove any variables that are not important to your analysis
  + Ensure missing values are coded as such (i.e. as `NA` instead of 0 or -1 or "missing")
  + Filling in missing values where you know what the answer should be (i.e. if a missing value really means 0 instead of missing)

## Example: Factor Levels in Wrong Order

```{r}
ames |> pull(Overall_Cond) |> levels()
```

## Re-Factoring

```{r}
ames <- ames |> 
  mutate(Overall_Cond = factor(Overall_Cond, levels = c("Very_Poor", "Poor", 
                                                        "Fair", "Below_Average",
                                                        "Average", "Above_Average", 
                                                        "Good", "Very_Good",
                                                        "Excellent", "Very_Excellent")))
ames |> pull(Overall_Cond) |> levels()
```

## Initial Data Split

```{r}
set.seed(427)

data_split <- initial_split(ames, strata = "Sale_Price")
ames_train <- training(data_split)
ames_test  <- testing(data_split)
```

## Zero-Variance (zv) and/or Near-Zero Variance (nzv) Variables {.smaller}

Heuristic for detecting near-zero variance features is:

- The fraction of unique values over the sample size is low (say ≤ 10%).
- The ratio of the frequency of the most prevalent value to the frequency of the second most prevalent
value is large (say ≥ 20%).

```{r}
library(caret)

nearZeroVar(ames_train, saveMetrics = TRUE) |> kable()
```
