---
title: 'MATH 427: Workflow Sets, Feature Selection, and Regularization'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(janitor) # for contingency tables
library(ISLR2)
library(readODS)

tidymodels_prefer()

set.seed(427)
```


# Workflow Sets in R

## Data: Different Ames Housing Prices {.smaller}

Goal: Predict `Sale_Price`.

```{r}
ames <- read_rds("../data/AmesHousing.rds")
ames |> glimpse()
```


```{r}
#| echo: FALSE
ames <- ames |> 
  mutate(Garage_Type = if_else(Garage_Type == "No_Garage", NA, Garage_Type),
        rand_num2 = runif(881, 0, 1),
        Year_Built = if_else(rand_num2 < 0.05, NA, Year_Built)) |> 
  select(-rand_num2)
```

## Clean Data Set

```{r}
ames <- ames |> 
  mutate(Overall_Qual = factor(Overall_Qual, levels = c("Very_Poor", "Poor", 
                                                        "Fair", "Below_Average",
                                                        "Average", "Above_Average", 
                                                        "Good", "Very_Good",
                                                        "Excellent", "Very_Excellent")),
         Garage_Type = if_else(is.na(Garage_Type), "No_Garage", Garage_Type),
         Garage_Type = as_factor(Garage_Type)
         )
```

## Initial Data Split

```{r}
data_split <- initial_split(ames, strata = "Sale_Price")
ames_train <- training(data_split)
ames_test  <- testing(data_split)
```

## Define Folds

```{r}
ames_folds <- vfold_cv(ames_train, v = 10, repeats = 10)
ames_folds
```

## Define Model(s)

```{r}
lm_model <- linear_reg() |> 
  set_engine("lm")

knn5_model <- nearest_neighbor(neighbors = 5) |>
  set_engine("kknn") |>
  set_mode("regression")

knn10_model <- nearest_neighbor(neighbors = 10) |>
  set_engine("kknn") |>
  set_mode("regression")
```

## Define Preprocessing: Linear regression

```{r}
lm_knnimpute <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_knn(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>  # in general use one_hot unless doing linear regression
  step_corr(all_numeric_predictors(), threshold = 0.5) |> # remove highly correlated predictors
  step_lincomb(all_numeric_predictors()) # remove variables that have exact linear combinations
```

## Define Preprocessing: Linear regression

```{r}
lm_meanimpute <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_mean(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>  # in general use one_hot unless doing linear regression
  step_corr(all_numeric_predictors(), threshold = 0.5) |> # remove highly correlated predictors
  step_lincomb(all_numeric_predictors()) # remove variables that have exact linear combinations
```

## Define Preprocessing: Linear regression

```{r}
lm_medianimpute <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_median(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>  # in general use one_hot unless doing linear regression
  step_corr(all_numeric_predictors(), threshold = 0.5) |> # remove highly correlated predictors
  step_lincomb(all_numeric_predictors()) # remove variables that have exact linear combinations
```

## Define Preprocessing: KNN

```{r}
knn_preproc1 <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_knn(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>  # in general use one_hot unless doing linear regression
  step_nzv(all_predictors()) |> 
  step_normalize(all_numeric_predictors())
```

## Define Preprocessing: KNN

```{r}
knn_preproc2 <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_mean(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>  # in general use one_hot unless doing linear regression
  step_nzv(all_predictors()) |> 
  step_normalize(all_numeric_predictors())
```

## Define Preprocessing: KNN

```{r}
knn_preproc3 <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_median(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>  # in general use one_hot unless doing linear regression
  step_nzv(all_predictors()) |> 
  step_normalize(all_numeric_predictors())
```

## Workflow Sets

-   Input `list`s of models and `recipe`s
-   If `cross = TRUE` will try out all combinations

## Create lists

```{r}
knn_preprocessors <- list(
  knn_knn_impute = knn_preproc1,
  knn_mean_impute = knn_preproc2,
  knn_median_imput = knn_preproc3
)

knn_models <- list(
  knn5 = knn5_model,
  knn10 = knn10_model
)
```

## Create lists

```{r}
lm_preprocessors <- list(
  lm_knn_impute = lm_knnimpute,
  lm_mean_impute = lm_meanimpute,
  lm_median_imput = lm_medianimpute
)

lm_models <- list(
  lm_model = lm_model
)
```

## Define Workflow Sets {.smaller}

```{r}
knn_models <- workflow_set(knn_preprocessors, knn_models, cross = TRUE)
lm_models <-  workflow_set(lm_preprocessors, lm_models, cross = TRUE)
all_models <- lm_models |> 
  bind_rows(knn_models)
  
all_models
```

## Define Metrics

```{r}
ames_metrics <- metric_set(rmse, rsq)
```

## Fit Resamples {.smaller}

```{r}
all_fits <- all_models |> 
  workflow_map("fit_resamples",
               resamples = ames_folds,
               metrics = ames_metrics)
```

## View Metrics

```{r}
collect_metrics(all_fits) |> 
  filter(.metric == "rmse") |> 
  kable()
```

## View Metrics

```{r}
collect_metrics(all_fits) |> 
  filter(.metric == "rsq") |> 
  kable()
```

## Plotting Results

```{r}
library(ggrepel)
autoplot(all_fits, metric = "rmse") +
  geom_text_repel(aes(label = wflow_id), nudge_x = 1/8, nudge_y = 1/100) +
  theme(legend.position = "none")
```

## Plotting Results

```{r}
autoplot(all_fits, metric = "rsq") +
  geom_text_repel(aes(label = wflow_id), nudge_x = 1/8, nudge_y = 1/100) +
  theme(legend.position = "none")
```




# Feature Selection for Linear Regression

## What is feature selection? {.smaller}

-   How do we choose what variables to include in our model?
-   Up to now... include all of them... probably not the best
-   Advantage of linear regression: interpretability
-   Including every feature decreases interpretability
-   Reasons for feature selection:
    +   Improve model performance
    +   Improve model interpretability
-   Parsimony: simpler models are called more *parsimonious*
-   Occam's Razor: more parsimonious models are better than less parsimonious models, holding all else constant

## Types of feature selection

-   Subset selection: Forward/Backward/Best-Subset Selection
-   Shrinkage-based methods: LASSO and Ridge Regression
-   Dimension reduction: consider linear combinations of predictors

# Subset Selection

## Exercise

-   With your group, write out the steps for the following algorithms on the board
  +   Group 1: Forward selection
  +   Group 2: Backward elimination
  +   Group 3: Step-wise selection
  +   Group 4: Best-subset selection

## Subset Selection in R {.smaller}

-   `tidymodels` does not have an implementation for any subset selection techniques
-   regularization (shrinkage-based) methods almost always perform better
-   [`colino` package](https://stevenpawley.github.io/colino/) provides `tidymodels` implementation
-   Other options
    +   `caret` package
    +   `olsrr` and `blorr` packages if you don't care about cross-validation
    +   implement yourself
    
## Feature Selection in R

-   When creating your recipe, don't need to always include all variables in your `recipe`:

```{r}
#| eval: FALSE
int_recipe <- recipe(pred ~ var1 + var2 + var1*var2, data = training_data) |> 
  step_x(...)
```

## Re-using Recipe but changing formula

```{r}
#| eval: FALSE
noint_recipe2 <- new_recipe |> 
  remove_formula() |> 
  add_formula(pred ~ var1 + var2)
```

# Shrinkage/Penalized/Regularization Methods

## Question

-   What criteria do we use to fit a linear regression model? Write down an equation with $\hat{\beta_i}$'s, $x_{ij}$'s, and $y_j$'s in it.

## OLS {.smaller}

-   Ordinary Least Squares Regression:

$$
\begin{aligned}
\hat{\beta} =\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}SSE(\hat{\beta}) &= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\sum_{j=1}^n(y_j-\hat{y}_j)^2\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2
\end{aligned}
$$

-   $\hat{\beta} = (\hat{\beta}_0,\ldots, \hat{\beta}_p)$ is the vector of all my coefficients
-   $\operatorname{argmin}$ is a function (operator) that returns the *arguments* that minimize the quantity it's being applied to

## Ridge Regression {.smaller}

$$
\begin{aligned}
\hat{\beta} &=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}} \left(SSE(\hat{\beta}) + \lambda\|\hat{\beta}\|_2^2\right) \\
&= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{y}_j)^2 + \lambda\sum_{i=1}^p \hat{\beta}_i^2\right)\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2 + \lambda\sum_{i=1}^p \hat{\beta}_i^2\right)
\end{aligned}
$$

-   $\lambda$ is called a "tuning parameter" and is not estimated from the data (more on this later)
-   Idea: penalize large coefficients
-   If we penalize large coefficients, what's going to happen to to our estimated coefficients? [[THEY SHRINK!]{.span style="color:red;"}]{.fragment .fade-in}
-   $\|\cdot\|_2$ is called the *$L_2$-norm*

## But... but... but why?!? {.smaller}

::: incremental
-   Recall the Bias-Variance Trade-Off
-   Our reducible error can partitioned into:
    +   Bias: how much $\hat{f}$ misses $f$ by *on average*
    +   Variance: how much $\hat{f}$ moves around from sample to sample
-   Ridge: increase bias a little bit in exchange for large decrease in variance
-   As we increase $\lambda$ do we increase or decrease the penalty for large coefficients?
-   As we increase $\lambda$ do we increase or decrease the *flexibility* of our model?
:::

## LASSO {.smaller}

$$
\begin{aligned}
\hat{\beta} &=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}} \left(SSE(\hat{\beta}) + \lambda\|\hat{\beta}\|_1\right) \\
&= \underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{y}_j)^2 + \lambda\sum_{i=1}^p |\hat{\beta}_i|\right)\\
&=\underset{\hat{\beta}_0,\ldots, \hat{\beta}_p}{\operatorname{argmin}}\left(\sum_{j=1}^n(y_j-\hat{\beta}_0-\hat{\beta}_1x_{1j} - \cdots - \hat{\beta}_px_{pj})^2 + \lambda\sum_{i=1}^p |\hat{\beta}_i|\right)
\end{aligned}
$$

-   $\lambda$ is called a "tuning parameter" and is not estimated from the data (more on this later)
-   Idea: penalize large coefficients
-   If we penalize large coefficients, what's going to happen to to our estimated coefficients [THEY SHRINK!]{.span style="color:red;"}
-   $\|\cdot\|_1$ is called the *$L_1$-norm*

## Question

-   What should happen to our coefficients as we *increase* $\lambda$?

## Ridge vs. LASSO

::: incremental
-   Ridge has a *closed-form solution*... how might be calculate it?
-   Ridge has some nice linear algebra properties that makes it EXTREMELY FAST to fit
-   LASSO has no *closed-form solution*... why?
-   LASSO coefficients estimated *numerically*... how?
    +   Gradient descent works (kind of) but something called **coordinate descent** is typically better
-   MOST IMPORTANT PROPERTY OF LASSO: it *induces sparsity* while Ridge does not
:::

## Sparsity in Applied Math {.smaller}

-   **sparse** typically means "most things are zero"
-   Example: sparse matrices are matrices where most entries are zero
    +   for large matrices this can provide HUGE performance gains
-   LASSO *induces sparsity* by setting most of the parameter estimates to zero
    +   this means it fits the model and does feature selection SIMULTANEOUSLY
-   Let's do some board work to see why this is...

## LASSO and Ridge in R

```{r}
ols <- linear_reg() |> 
  set_engine("lm")
ridge_0 <- linear_reg(mixture = 0, penalty = 0) |> # penalty set's our lambda
  set_engine("glmnet")
ridge_1000 <- linear_reg(mixture = 0, penalty = 1000) |> # penalty set's our lambda
  set_engine("glmnet")
ridge_10000 <- linear_reg(mixture = 0, penalty = 10000) |> # penalty set's our lambda
  set_engine("glmnet")
lasso_0 <- linear_reg(mixture = 1, penalty = 0) |> # penalty set's our lambda
  set_engine("glmnet")
lasso_1000 <- linear_reg(mixture = 1, penalty = 1000) |> # penalty set's our lambda
  set_engine("glmnet")
lasso_10000 <- linear_reg(mixture = 1, penalty = 10000) |> # penalty set's our lambda
  set_engine("glmnet")
```

## Create Recipe

```{r}
lm_meanimpute <- recipe(Sale_Price ~ ., data = ames_train) |> 
  step_nzv(all_predictors()) |>  # remove zero or near-zero variable predictors
  step_impute_mean(Year_Built, Gr_Liv_Area) |>  # impute missing values in Overall_Qual and Year_Built
  step_integer(Overall_Qual) |> # convert Overall_Qual into ordinal encoding
  step_other(all_nominal_predictors(), threshold = 0.01, other = "Other") |> # lump all categories with less than 1% representation into a category called Other for each variable
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>  # in general use one_hot unless doing linear regression
  step_normalize(all_numeric_predictors()) |> 
  step_corr(all_numeric_predictors(), threshold = 0.5) |> # remove highly correlated predictors
  step_lincomb(all_numeric_predictors()) # remove variables that have exact linear combinations
```

## Question

-   Why do we need to normalize our data?

## Create workflows

```{r}
generic_wf <- workflow() |> add_recipe(lm_meanimpute)
ols_wf <- generic_wf |> add_model(ols)
ridge0_wf <- generic_wf |> add_model(ridge_0)
ridge1000_wf <- generic_wf |> add_model(ridge_1000)
ridge10000_wf <- generic_wf |> add_model(ridge_10000)
lasso0_wf <- generic_wf |> add_model(lasso_0)
lasso1000_wf <- generic_wf |> add_model(lasso_1000)
lasso10000_wf <- generic_wf |> add_model(lasso_10000)
```

## Fit Models

```{r}
ols_fit <- ols_wf |> fit(ames_train)
ridge0_fit <- ridge0_wf |> fit(ames_train)
ridge1000_fit <- ridge1000_wf |> fit(ames_train)
ridge10000_fit <- ridge10000_wf |> fit(ames_train)
lasso0_fit <- lasso0_wf |> fit(ames_train)
lasso1000_fit <- lasso1000_wf |> fit(ames_train)
lasso10000_fit <- lasso10000_wf |> fit(ames_train)
```

## Collect coefficients

```{r}
all_coefs <- bind_cols(model = "ols", tidy(ols_fit)) |> 
  bind_rows(bind_cols(model = "ridge0", tidy(ridge0_fit))) |> 
  bind_rows(bind_cols(model = "ridge1000", tidy(ridge1000_fit))) |> 
  bind_rows(bind_cols(model = "ridge10000", tidy(ridge10000_fit))) |> 
  bind_rows(bind_cols(model = "lasso0", tidy(lasso0_fit))) |> 
  bind_rows(bind_cols(model = "lasso1000", tidy(lasso1000_fit))) |> 
  bind_rows(bind_cols(model = "lasso10000", tidy(lasso10000_fit)))
```

## Question

::: incremental
-   What should we expect from `ols`, `ridge0`, and `lasso0`?
-   They should be the same!
:::

## Visualize

```{r}
all_coefs |> 
  filter(model %in% c("ols", "ridge0", "lasso0")) |> 
  ggplot(aes(y = term, x = estimate, fill = model, color = model)) +
  geom_bar(position = "dodge", stat = "identity")
```

## Uh-Oh!

::: incremental
-   They're not the same
-   Any idea why?
-   Algorithm used to fit model
    +   `lm` estimates coefficients *analytically*
    +   `glmnet` estimates coefficients *numerically* using an algorithm named "coordinate-descent"
-   Moral: you must understand theory AND application
:::

## Visualizing Coefficients: Ridge

```{r}
all_coefs |> 
  filter(model %in% c("ols", "ridge1000", "ridge10000")) |> 
  ggplot(aes(y = term, x = estimate, fill = model, color = model)) +
  geom_bar(position = "dodge", stat = "identity")
```

## Visualizing Coefficients: LASSO

```{r}
all_coefs |> 
  filter(model %in% c("ols", "lasso1000", "lasso10000")) |> 
  ggplot(aes(y = term, x = estimate, fill = model, color = model)) +
  geom_bar(position = "dodge", stat = "identity")
```

## Next Time

-   How do we choose $\lambda$?