---
title: 'MATH 427: Model Tuning'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(fivethirtyeight)
library(knitr)
tidymodels_prefer()

set.seed(427)
```


## Data: Candy {.smaller}

The data from this lab comes from the the article FiveThirtyEight [*The Ultimate Halloween Candy Power Ranking*](https://fivethirtyeight.com/features/the-ultimate-halloween-candy-power-ranking) by Walt Hickey. To collect data, Hickey and collaborators at FiveThirtyEight set up an experiment people could vote on a series of randomly generated candy matchups (e.g. Reese's vs. Skittles). Click [here](http://walthickey.com/2017/10/18/whats-the-best-halloween-candy/) to check out some of the match ups.

The data set contains 12 characteristics and win percentage from 85 candies in the experiment.

## Data: Candy

```{r}
glimpse(candy_rankings)
```

## Data Cleaning

```{r}
candy_rankings_clean <- candy_rankings |> 
  select(-competitorname) |> 
  mutate(sugarpercent = sugarpercent*100,
         pricepercent = pricepercent*100)
```

## Data Splitting

```{r}
candy_split <- initial_split(candy_rankings_clean, strata = winpercent)
candy_train <- training(candy_split)
candy_test <- testing(candy_split)
```

## Pre-processing

```{r}
lm_preproc <- recipe(winpercent ~ . , data = candy_train) |> 
  step_nzv(all_predictors()) |> 
  step_integer(all_logical_predictors()) |> 
  step_dummy(all_nominal_predictors())

lm_preproc |> prep()
```