---
title: 'MATH 427: Classification Trees'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---


## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(dsbox) # dcbikeshare data
library(knitr)

tidymodels_prefer()

set.seed(427)
```



## Decision Trees

-   Advantages
    +   Easy to explain and interpret
    +   Closely mirror human decision-making
    +   Can be displayed graphically, and are easily interpreted by non-experts
    +   Does not require standardization of predictors
    +   Can handle missing data directly
    +   Can easily capture non-linear patterns
-   Disadvantages
    +   Do not have same level of prediction accuracy
    +   Not very robust
    
# Decision Trees for Classification

## Last Time

::: incremental
-   Regression Trees: Decision Trees for Regression Problems
-   How are they fit?
-   What is pruning? Why do we do it?
-   What tuning parameter did we talk about last time?
-   Today: Classification Trees
:::

## Classification Trees

-   Predictions: 
    +   Classes: most common class at terminal node
    +   Probability: proportion of each class at terminal node
-   Rest of tree: same as regression tree

## Fitting Classification Trees {.smaller}
-   Still use **recursive binary splitting** to grow a classification tree
-   $\hat{p}_{mk}$: proportion of training observations in the $m^{th}$ region from the $k^{th}$ class
-   $SSE$ can be replaced by 
    +   **classification error rate**, fraction of the training observations that do not belong to the most common class ($1 - \text{accuracy}) $$E = 1 - \max_k \left(\hat{p}_{mk}\right)$$
    +   **Gini index**,  measure of node purity— small valuse indicate that a node is predominantly a single class $$G = \displaystyle \sum_{k=1}^{K} \hat{p}_{mk}\left(1-\hat{p}_{mk}\right)$$

## Data: Voter Frequency

[Info about data](https://github.com/fivethirtyeight/data/tree/master/non-voters)

```{r}
voter_data <- read_csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv')

glimpse(voter_data)
```

## Cleaning Data: Missing Data

```{r}
voter_data |> 
  summarize(across(everything(), ~sum(is.na(.x)))) |> 
  pivot_longer(everything()) |> 
  filter(value > 0)
```

## Cleaning the Data: Q28, 29, 31

-   What should we do with question 28?
-   What should we do with question 29?
-   What should we do with question 31?

## Cleaning the Data

```{r}
voter_clean <- voter_data |> 
  select(-RespId, -weight) |> 
  mutate(across(Q28_1:Q29_10, ~as_factor(case_when(
    .x == 1 ~ "Yes",
    .x == -1 ~ "No",
    TRUE ~ "Never Asked"
  ))),
  Party_ID = as_factor(case_when(
    Q31 == 1 ~ "Strong Republican",
    Q31 == 2 ~ "Republican",
    Q32 == 1  ~ "Strong Democrat",
    Q32 == 2 ~ "Democrat",
    Q33 == 1 ~ "Lean Republican",
    Q33 == 2 ~ "Lean Democrat"
  )))
```