---
title: 'MATH 427: Classification Trees'
author: Eric Friedlander
footer: "[🔗 MAT 427 - Spring 2025 -  Schedule](https://mat427sp25.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  cache: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---


## Computational Set-Up

```{r}
library(tidyverse)
library(tidymodels)
library(dsbox) # dcbikeshare data
library(knitr)
library(kableExtra)

tidymodels_prefer()

set.seed(427)
```



# Decision Trees for Classification

## Classification Trees

-   Predictions: 
    +   Classes: most common class at terminal node
    +   Probability: proportion of each class at terminal node
-   Rest of tree: same as regression tree

## Classification Trees

-   Predictions: 
    +   Classes: most common class at terminal node
    +   Probability: proportion of each class at terminal node
-   Rest of tree: same as regression tree

## Fitting Classification Trees {.smaller}
-   Still use **recursive binary splitting** to grow a classification tree
-   $\hat{p}_{mk}$: proportion of training observations in the $m^{th}$ region from the $k^{th}$ class
-   $SSE$ can be replaced by 
    +   **classification error rate**, fraction of the training observations that do not belong to the most common class ($1 - \text{accuracy}) $$E = 1 - \max_k \left(\hat{p}_{mk}\right)$$
    +   **Gini index**,  measure of node purity— small valuse indicate that a node is predominantly a single class $$G = \displaystyle \sum_{k=1}^{K} \hat{p}_{mk}\left(1-\hat{p}_{mk}\right)$$
    
## General Idea

-   Many different implementations so check documentation
-   ISLR2 recommendation
    +   Use Gini or Cross-Entropy to split
    +   Use classification error rate to prune
-   `rpart` uses Gini to split and classification error (plus penalty) to prune

# Data Cleaning Side-Quest

## Data: Voter Frequency

-   [Info about data](https://github.com/fivethirtyeight/data/tree/master/non-voters)
-   Goal: Identify individuals who are unlikely to vote to help organization target "get out the vote" effort.

```{r}
voter_data <- read_csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv')

glimpse(voter_data)
```

## Cleaning the Data: Straight Forward Stuff

```{r}
voter_clean <- voter_data |> 
  select(-RespId, -weight, -Q1) |>
  mutate(
    educ = factor(educ, levels = c("High school or less", "Some college", "College")),
    income_cat = factor(income_cat, levels = c("Less than $40k", "$40-75k ",
                                               "$75-125k", "$125k or more")),
    voter_category = factor(voter_category, levels = c("rarely/never", "sporadic", "always"))
  )
```

## Cleaning Data: Missing Data

```{r}
voter_data |> 
  summarize(across(everything(), ~sum(is.na(.x)))) |> 
  pivot_longer(everything()) |> 
  filter(value > 0) |> 
  kable() |> 
  kable_paper() |> 
  scroll_box(width = "1050px", height = "400px")
```

## Cleaning the Data: Q22, 28, 29, 31, 32, 33

-   Dealing with missing values should depend on what "missing" means in the context of each variable

## Cleaning the Data: Q22

```{r}
voter_clean |> 
  count(Q22) |> 
  kable() |> 
  kable_paper() |> 
  scroll_box(width = "1050px", height = "400px")
```

## Uh-Oh... -1s?

:::: columns
::: column
```{r}
voter_clean |> filter(Q22 == -1) |> select(Q20, Q22) |> kable() |> 
  kable_paper() |> scroll_box(width = "1050px", height = "400px")
```
:::

::: column
```{r}
voter_clean |> filter(is.na(Q22)) |> select(Q20, Q22) |> kable() |> 
  kable_paper() |> scroll_box(width = "1050px", height = "400px")
```
:::
::::

## More Cleaning: -1s

```{r}
voter_clean |> 
  summarize(across(everything(), ~sum(.x == -1))) |> 
  pivot_longer(everything()) |> 
  filter(value > 0)  |> kable() |> 
  kable_paper() |> scroll_box(width = "1050px", height = "400px")
```

## How should we handle -1s? {.smaller}

::: incremental
-   Seems like -1 means: question asked and not answered, or answer not ranked
-   Seems like `NA` means: question never asked
-   What should we do?
-   Couple methods:
    +   Replace -1 with "Not answered"
    +   Replace -1 with `NA`
    +   Replace -1 with `NA` and add extra columns `QX_Ranked_Answered`, (Yes/No)
        + Can do this using `step_indicate_na` in our recipe
        + First, deal with current `NA`s
        + Second, convert -1s to `NA`
:::

## Dealing with current `NA`s: Q22

```{r}
voter_clean <- voter_clean |> 
  filter(Q22 != 5) |> 
  mutate(Q22 = as_factor(Q22),
         Q22 = if_else(is.na(Q22), "Not Asked", Q22))
```

## Cleaning Q28

```{r}
voter_clean |> 
  count(Q28_1) |> 
  kable()
```

## Cleaning Q28

-   Seems like 
    +   -1 means not selected
    +   Seems like 1 means selected
    +   `NA` means never asked

## Cleaning Q28

```{r}
voter_clean <- voter_clean |> 
  mutate(across(Q28_1:Q28_8, ~if_else(.x == -1, 0, .x)),
         across(Q28_1:Q28_8, ~ as_factor(.x)),
         across(Q28_1:Q28_8, ~if_else(is.na(.x) , "Not Asked", .x))
  )
```

## Cleaning Q29

```{r}
voter_clean <- voter_clean |> 
  mutate(across(Q29_1:Q29_10, ~if_else(.x == -1, 0, .x)),
         across(Q29_1:Q29_10, ~ as_factor(.x)),
         across(Q29_1:Q29_8, ~if_else(is.na(.x) , "Not Asked", .x))
  )
```


## Cleaning Q31:Q33: Handling Party

```{r}
voter_clean <- voter_clean |> 
mutate(
  Party_ID = as_factor(case_when(
    Q31 == 1 ~ "Strong Republican",
    Q31 == 2 ~ "Republican",
    Q32 == 1  ~ "Strong Democrat",
    Q32 == 2 ~ "Democrat",
    Q33 == 1 ~ "Lean Republican",
    Q33 == 2 ~ "Lean Democrat",
    TRUE ~ "Other"
  )),
  Party_ID = factor(Party_ID, levels =c("Strong Republican", "Republican", "Lean Republican",
                                          "Other", "Lean Democrat", "Democrat", "Strong Democrat")))
```

## Finally: Dealing with -1s

```{r}
voter_clean <- voter_clean |> 
  mutate(across(!ppage, ~as_factor(if_else(.x == -1, NA, .x))))
```

# Classification Trees in R

## Split Data

```{r}
set.seed(427)

voter_splits <- initial_split(voter_clean, prop = 0.7, strata = voter_category)
voter_train <- training(voter_splits)
voter_test <- testing(voter_splits)
```

## Define Model

```{r}
tree_model <- decision_tree(cost_complexity = tune()) |> 
  set_engine("rpart") |> 
  set_mode("classification")
```

## Define Recipe

```{r}
tree_recipe <- recipe(voter_category ~ . , data = voter_train) |> 
  step_indicate_na(all_predictors()) |> 
  step_integer(educ, income_cat, Party_ID, Q2_2:Q4_6, Q6, Q8_1:Q9_4, Q14:Q17_4,
               Q25:Q26) |> 
  step_impute_median(all_numeric_predictors()) |> 
  step_impute_mode(all_nominal_predictors()) |> 
  step_dummy(all_nominal_predictors(), one_hot = TRUE)
```

## Define Workflow

```{r}
tree_wf <- workflow() |> 
  add_model(tree_model) |> 
  add_recipe(tree_recipe)
```

## Tune Model

```{r}
voter_folds <- vfold_cv(voter_train, v = 2, repeats = 2)
cp_grid <- grid_regular(cost_complexity(range = c(-10, 0)), levels = 10)
tuned_model <- tune_grid(tree_wf,
                         resamples = voter_folds,
                         grid = cp_grid)
```

# Plot output

```{r}
autoplot(tune_res)
```

