---
title: "Homework 3: $K$-Nearest Neighbors"
author: "Your Name"
editor: visual
format:
  html:
    embed-resources: true
---

```{r setup}
#| include: false

# load packages here
```

# Introduction

In this homework, you will practice applying the $K$-Nearest Neighbors (KNN) method which is capable of performing both classification and regression. You will also practice collaborating with a team over GitHub.

## Learning goals

By the end of the homework, you will...

-   Be able to work simultaneously with teammates on the same document using GitHub
-   Fit and interpret linear models
-   Split data using `tidymodels`
-   Compare and evaluate different linear models

## $K$-Nearest Neighbors

The basic idea is that predictions are made based on the $K$ observation in the training data which are "closest" to the observation that we're making a prediction for. While many different **metrics** (i.e. measures of distance) can be used, we will work exclusively with the **Euclidean metric**: $$\text{dist}(x, y) = \sqrt{\sum_{i=1}^p(x_i-y_i)^2}$$ for vectors $x = (x_1,\ldots, x_p)$ and $y = (y_1,\ldots,y_p)$.

# Getting Started

In last weeks homework you learned how to share your work using GitHub but only while working on the same document at different times. This week we will learn how to work on the same document *simultaneously*. 

## Teams

You can find your team for this assignment on Canvas in the **People** section. The group set is called **HW3**. Your group will consist of 2-3 people and has been randomly generated. But first, some rules:

1.  You and your team members should be in the same physical room when you complete this assignment.
2.  You should all contribute to each problem, but to receive credit you must rotate who actually writes down the answers.
3.  In order to receive credit, you must have a commit after each exercise by the correct member of your team.
4.  For now, don't try to edit the same document at the same time. We will cover that in later homeworks.

## Clone the repo & start new RStudio project

The following directions will guide you through the process of setting up your homework to work as a group.

## Exercise 0.1

::: {.callout-tip title="Question"}
In your group, decide on a team name. Then have **one member of your group**:

1.  [Click this link](https://classroom.github.com/a/3cnqO-bh) to accept the assignment and enter your team name.
2.  Repeat the directions for creating a project from HW 1 with the HW 3 repository.

Once this is complete, the other members can do the same thing, being careful to join the already created team on GitHub classroom.
:::

## Exercise 0.2

We will now learn how to collaborate on the same document at the same time.

::: {.callout-tip title="Question"}
Have Member 1 fill in the Team Name and their name below and have Members 2  & 3 fill in their names. If you only have two members Member 2 should delete the line for Member 3. 
:::

-   Team Name: \[Insert Team Name\]
-   Member 1: \[Insert Name\]
-   Member 2: \[Insert Name\]
-   Member 3: \[Insert Name/Delete line if you only have two members\]

## KNN for Classification

We'll start by using KNN for classification.

## Data

We will be working with the famous `iris` data set which consists of four measurements (in centimeters) for 150 plants belonging to three species of iris. This data set was first published in a classic 1936 paper by English statistician, and notable racist/proponent of Eugenics, Ronald Fisher. In that paper, multivariate linear models were applied to classify these plants. Of course, back then, model fitting was an extremely laborious process that was done without the aid of calculators or statistical software. To access this data first load the package `datasets` and then load the data set `iris`.

## Exercise 1

::: {.callout-tip title="Question"}
Import the `iris` data set from the `datasets` package and take a look at the columns. We would like to visualize the $K$-nearest neighbor method in two dimensions, so make a data set called `iris2` consisting of the columns `Sepal.Width`, `Petal.Width`, and `Species`. Split your data into training and test sets using a 70-30 split. IMPORTANT: Make sure that each species is represented proportionally in the training set by using the `strata` argument in the `initial_split` function! Once again, set your seed to 301.
:::

## The KNN algorithm

## Exercise 2

::: {.callout-tip title="Question"}
Plot the points in your training set in the xy-plane colored by the `Species` labels.
:::

## Exercise 3

::: {.callout-tip title="Question"}
As the name suggests, the $K$-nearest neighbors (KNN) method classifies a point based on the classification of the observations in the training set that are nearest to that point. If $k > 1$, then the neighbors essentially "vote" on the classification of the point. Based on your graph, if $k = 1$, how would KNN classify a flower that had sepal width 3cm and petal width 1cm?
:::

## Exercise 4

::: {.callout-tip title="Question"}
Just to verify that we are correct, find the sepal width, petal width, and species of the observation in your training set that is closest to our flower with sepal width 3cm and petal width 1cm.
:::

## Exercise 5

::: {.callout-tip title="Question"}
Center and scale your data sets. You should be using the mean and standard deviation from your training set to center and scale both data sets.
:::

## Exercise 6

::: {.callout-tip title="Question"}
Rather than implementing this method by hand, we can use the function `knn3` in the `caret` package.
:::

We would like to understand how the method of $K$-nearest neighbors will classify points in the plane. That is, we would like to view the *decision boundaries* of this model. To do this, we will use our model to classify a large grid of points in the plane, and color them by their classification. The code below creates a data frame called `grid` consisting of `r 250^2` points in the plane.

```{r}
# g1 <- rep((200:450)*(1/100), 250)
# g2 <- rep((0:250)*(1/100), each = 250)
# grid <- tibble(  x1 = g1
#                    , x2 = g2)
```

## Exercise 7

::: {.callout-tip title="Question"}
Uncomment the code above and classify the points in `grid` using your training data and $k = 1$. Then, plot the points in `grid` colored by their classification. Don't forget to center and scale the grid.
:::

## Exercise 8

::: {.callout-tip title="Question"}
Notice that the decision boundary between versicolor and virginica looks a little strange. What do you observe? Why do you think this is happening? Does using $k = 2$ make things better or worse? Why do you think that is?
:::

## Exercise 9

::: {.callout-tip title="Question"}
Determine which value of $K$, the number of neighbors selected, gives the highest accuracy on the test set. Test all $K$s between 1 and 40. Note that there may be ties because our data set is a little bit too small. To break ties just choose the smallest $K$ among the ones which are tied.
:::

## Exercise 10

::: {.callout-tip title="Question"}
Use your value of $K$ to classify the points in `test` based on this combined data set. You must recenter and scale your data based on this combination. Generate a confusion matrix and report the accuracy of your method? *Note.* You should "set the seed" in R by specifying `set.seed(301)` before you build your model. Because R randomly breaks ties, if you do not set the seed, you may get a different result the next time you knit your document (and your answer won't match your code).

Awesome!! Your model probably did pretty well, because KNN performs really well on the `iris` data set. However, this isn't a very challenging data set for most classification methods. More challenging data sets have data on different scales and *class imbalance* where there are very few observations belonging to a particular class.
:::

## Exercise 11

::: {.callout-tip title="Question"}
KNN can also be modified to give probabilistic predictions like we get from logistic regression. How would you modify the method to give a probabilistic prediction instead of a classification?
:::

# KNN for Regression

For regression, we can predict the response variable for our point to be the average (or sometimes median) of the response variable for the $K$-nearest neighbors.

## Data & Dummy Variables

For this portion of the homework, we'll return to using the `Carseat` data from the `ISLR` package that we worked with in homework 4. Frequently, when working with categorical data, you will be required to transform that data into **dummy variables**. Namely, you'll create a unique variable for each column which gets a 1 if the corresponding observation is from that category and a 0 otherwise. In data science, this format is sometimes referred to as **one-hot encoding**.

## Exercise 12

::: {.callout-tip title="Question"}
Load the `Carseat` data from the `ISLR` package. Use the `dummyVars` function from the `caret` package to create dummy variables for the categorical variables in `Carseat`. Then, split the data into a training, validation, and test set using a 60-20-20 split and a seed of 301 (as usual). Explain what the `fullRank` options does and why we can create dummy variables before splitting our data.
:::

There is no standard way of handling categorical variables when applying KNN to predictors that are categorical. This is because we need to answer the fundamental question of "What is the distance between two observations that belong to different categories?". There are several ways of approaching this. One is to use different metrics and another is to just convert your categorical variables into numerical variables by using this one-hot encoding. Past that, you'll need to decide whether to center and standardize your dummy variables.

## Using KNN for Regression

The function `knnreg` from the `caret` package will apply KNN to a regression problem by taking the average of the $K$ nearest neighbors to the observation you're looking to make a prediction for. `knnreg` works very similar to `knn3` and previous models that we've fit in R.

## Exercise 13

::: {.callout-tip title="Question"}
Fit a KNN model to predict `Sales` from the data we have. Fit your model on the training data and use the validation set to choose the appropriate variable and the number of neighbors to include. You may find it useful to plot the $R^2$ and RMSE against the number of neighbors you include in your model. Once you select a $K$ and the variables you want to include in your model, see how it performs on the test set and compare your result to the linear model you fit at the end of homework 4. Which model works better? You will find that the RMSE and $R^2$ disagree on what the best model is. You will have to make a judgement call on which model is "best". One thing that can be helpful is looking at plots of your target variables (`Sales` in this case) against the model residuals. When you do this, what pattern do you see?
:::
